---
title: "Class 9: Candy Mini-Project"
author: "Daniel Kim"
format: pdf
---

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

9 different candy types. If the column is type int it is a candy type.

> Q2. How many fruity candy types are in the dataset?

```{r}
table(candy$fruity)
```

38 fruity candies are in this data set.

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["Snickers", ]$winpercent
```

My favorite candy is Snickers and it's winpercent value is 76.67378.

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

76.7686.

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

49.6535.

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

It seems like sugarpercent and pricepercent are decimal points in the range of 0-1, however, winpercent is actual percentage points in the range of 0-100. The reset are 0 or 1 binary.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

Zero means not a chocolate and one means is a chocolate.

> Q8. Plot a histogram of winpercent values using both base R an ggplot2.

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)
ggplot(candy, aes(x = winpercent)) + geom_histogram(bins = 10)
```

> Q9. Is the distribution of winpercent values symmetrical?

No, it is not symmetrical. It is slightly skewed right.

> Q10. Is the center of the distribution above or below 50%?

The center is below 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```

On average, chocolate candy is ranked higher than fruit candy.

> Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```

Yes, the difference is statistically significant. The p-value of 2.871e-08 is lower than 0.05 alpha.

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

The 5 least liked candies are in order, Nik L Nip, Boston Baked beans, Chiclets, Super Bubble, and Jawbusters.

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE),], n=5)
```

The 5 most liked candies are in order, Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers.

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) + 
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) + 
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starbursts.

```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(x = winpercent, y = pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
copy <- candy
copy$value <- copy$winpercent / copy$pricepercent
head(copy[order(copy$value, decreasing = TRUE),])
```

Tootsie Roll Midgies have the best value in terms of winpercent.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The top 5 most expensive candy types are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Kracekl, and Hershey's Milk Chocolate. Out of these Nik L Nip is the least popular.

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are anti-correlated.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent or bar are most psotively correlated.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
ggplot(pca$rotation) +
  aes(x=PC1, reorder(rownames(pca$rotation), PC1)) +
  geom_col()
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

Fruity, Pluribus, and Hard is picked up strongly by PC1 in the positive direction. They make sense, fruity candies like starbursts are hard and come in bags. This relationship was shown in the PC1 vs PC2 graph where the fruity candies were clustered on the right side - positive PC1 score.

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

There are couple of characteristics that make a candy "winning". The first one is Chocolate wins more than frutiy candies. Bar chocolates and winpercent are grouped together. And peanut flavor is also grouped together. 

