---
title: "HW 12: Population Scale Analysis"
author: "Daniel Kim"
format: pdf
---

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
file <- read.table("data.txt", header = TRUE)
```

```{r}
summary(file)
table(file$geno)
median(file$exp[file$geno == "A/A"])
median(file$exp[file$geno == "A/G"])
median(file$exp[file$geno == "G/G"])
```

Sample size of 108 for A/A, 233 for A/G and 121 for G/G. Median expression level of 31.24847 for A/A, 25.06486 for A/G, and 20.073673 for G/G.

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(file, aes(x = geno, y = exp, fill = geno)) +
  geom_boxplot() +
  geom_jitter() +
  labs(x = "Genotype", y = "Expression")
```

A/A shows the highest median expression levels and A/G lower, and G/G lowest. It seems like each additional G allele decreases the expression levels of the ORMDL3 gene. Yes it does affect the expression.

