---
title: "Class 6: R Functions"
author: "Daniel Kim"
format: pdf
---

This week we are introducing **R Functions** and how to write our own R functions.

```{r}
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

Questions to answer:

> Q1. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
grade <- function(vector) {
  #replaces all NAs with 0
  vector[is.na(vector)] <- 0
  
  #performs a average calculation on a updated vector in which the lowest value has been removed
  return (mean(vector[-which.min(vector)]))
}
```

> Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts]

```{r}
#reads csv from link, use first columns as name
csv <- read.csv("https://tinyurl.com/gradeinput", row.names=1)
#apply the grade function to csv rows
scores <- apply(csv, 1, grade)
#returns the max scorer and their index
which.max(scores)
```
> Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]

```{r}
#applies the mean function while removing NAs to csv columns
avg_scores <- apply(csv, 2, mean, na.rm=TRUE)
#returns which column aka homework had lowest average scores
which.min(avg_scores)
```
> Q4. Optional Extension: From your analysis of the gradebook, which homework was most
predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

```{r}
#creates a copy of csv
masked_csv <- csv
#replaces all NAs with 0 in the copied csv
masked_csv[is.na(masked_csv)] <- 0
#returns the index after applying cor function to csv on columns where x is scores
which.max(apply(masked_csv, 2, cor, x=scores))
```

