---
title: "Class 7: Machine Learning 1"
author: "Daniel Kim"
format: pdf
---

# Principal Component Analysis (PCA)

## PCA of UK food data

Read data from website and try a few visualizations.

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

The dim() function can be used to answer this question. There are 17 rows and 5 columns.

```{r}
dim(x)
head(x)
```

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
dim(x)
```

```{r}
x <- read.csv(url, row.names=1)
head(x)
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer using the second option, because, for the first one if it is ran multiple times, the row number will decrease by 1 each time. This is not good for code usability, and the second option is more robust. I can mistakenly delete the whole data frame.

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

> Q3: Changing what optional argument in the above barplot() function results in the following plot?

Changing the beside=T argument to blank makes the following change. Its default parameter is false so leaving it blank or setting it to false results the same thing.

```{r}
barplot(as.matrix(x), col=rainbow(nrow(x)))
```

```{r}
library(tidyr)

x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)
head(x_long)
```

```{r}
library(ggplot2)
```

```{r}
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_bw()
```

> Q4: Changing what optional argument in the above ggplot() code results in a stacked barplot figure?

Using the geom_col(position = "stacked") instead of geom_col(position = "dodge") argument will result in a stacked barplot figure.

```{r}
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(positon = "stacked") +
  theme_bw()
```

> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

Every graph is a graph of a country vs another country compared to each other. The first row is y=England, and starting from the second column its Wales, Scotland, N.Ireland... And it goes on for every row. If there is a point that lies in the diagonal for a given plot it is a data point that is very close to the correlation between the 2 countries being compared. If point x is on the diagonal, it means the point fits very well into their line of best fit.

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```

> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

Based on the heat map, England and Wales cluster together, it means their food consumption patterns are the most similar. N.Ireland is different because it consumes less Fresh Fruit and Other meat than the other countires, also have the highest consumption of fresh potatoes, and the lowest consumption of alcohol.

```{r}
pca <- prcomp( t(x) )
summary(pca)
```

> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

> Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

colors <- c("orange", "red", "blue", "green")

ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5, color = colors) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```

```{r}
z <- summary(pca)
z$importance
```

```{r}
variance_df <- data.frame(
  PC = factor(paste0("PC", 1:length(v)), levels = paste0("PC", 1:length(v))),
  Variance = v
)

ggplot(variance_df) +
  aes(x = PC, y = Variance) +
  geom_col(fill = "steelblue") +
  xlab("Principal Component") +
  ylab("Percent Variation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0))
```

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

The 2 most predominant food groups are alcoholic and soft drinks. PC2 mainly tells about the high consumption of soft drinks compared the the not so consumed fresh potatoes.